The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-03-14 products/manuscript/Manuscript.qmd
P Update

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Define the correct file path using the `here` package
data_path <- here("data", "raw-data", "PROMO_Data.csv")

# Read the CSV file, suppressing messages
PROMO_Data <- read_csv(data_path, show_col_types = FALSE)
```

@@

Malaria is a life-threatening disease caused by Plasmodium parasites, which are transmitted to humans through the bite of an infected female Anopheles mosquito (Talapko et al., 2019). According to the World Health Organization, in 2022, approximately 249 million malaria cases were reported across 85 endemic countries—an increase of 5 million cases compared to 2021. Notably, Uganda alone contributed more than 597,000 cases during that period (WHO, 2023). Despite significant efforts in prevention and treatment, malaria remains a major global health challenge. Vulnerable groups such as pregnant women, children, and immunocompromised individuals are at the highest risk, and Africa continues to bear the heaviest burden of this disease. Pregnant women with malaria tend to experience more severe clinical symptoms and poorer outcomes, including heightened risks of miscarriage, fetal loss, premature birth, low birth weight in newborns, and neonatal death (Caroline Lin Lin Chua et al., 2021). A multifaceted approach incorporating insecticide-treated bed nets (ITNs) and chemoprevention strategies has proven effective in mitigating the adverse health outcomes associated with malaria during pregnancy. Despite this progress, a 2017 report reveals that only 22% of pregnant women in Sub-Saharan Africa received an entire course of IPTp (Bauserman et al., 2019), underscoring a significant disparity in treatment access. Addressing this issue, the study seeks to: (a) examine the association between the frequency of malaria episodes during pregnancy and adverse birth outcomes, such as preterm birth and stillbirth; (b) assess whether infants born to mothers treated with DP are less likely to experience low birth weight compared to those treated with SP; and (c) explore the relationship between placental malaria and adverse birth outcomes, including preterm birth and low birth weight.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Data Import & Initial Inspection

# Read the dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"), show_col_types = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Data Cleaning & Transformation

# Load required libraries (if not already loaded)
library(janitor)   # For clean_names()

# Clean Column Names
PROMO_Data <- PROMO_Data %>% clean_names()

# Remove Empty Rows
PROMO_Data <- PROMO_Data %>%
  filter(!is.na(study_arm), !is.na(stillbirth), !is.na(neonatal_death), !is.na(infant_low_birth_weight_2500_g))

#  Handle Missing Values
PROMO_Data <- PROMO_Data %>%
  replace_na(list(
    study_arm = "Unknown",
    stillbirth = "Unknown",
    neonatal_death = "Unknown",
    placental_malaria = "Unknown",
    delivery_mode = "Unknown"
  )) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

#  Convert Data Types
PROMO_Data <- PROMO_Data %>%
  mutate(across(c(study_arm, stillbirth, neonatal_death, placental_malaria, delivery_mode), as.factor)) %>%
  mutate(across(contains("date"), as.Date, format="%Y-%m-%d"))

#  Verify Changes
glimpse(PROMO_Data)  # More readable than str()

#  Summary of Numeric Variables
print(skim(PROMO_Data))  # Explicitly print skim output

#  Frequency Counts of Key Categorical Variables
PROMO_Data %>% count(study_arm, sort = TRUE) %>% print()
PROMO_Data %>% count(placental_malaria, sort = TRUE) %>% print()
PROMO_Data %>% count(preterm_birth_ontoneo, sort = TRUE) %>% print()

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Load required libraries
library(dplyr)
library(gtsummary)

#  Select key baseline variables (ensure they exist)
baseline_vars <- c("study_arm", "education_level_omrse", "gravidity", "parity", 
                   "placental_malaria", "preterm_birth_ontoneo", "stillbirth", 
                   "neonatal_death", "infant_low_birth_weight_2500_g")

#  Ensure only existing variables are used
baseline_vars <- intersect(baseline_vars, colnames(PROMO_Data))

# Generate table only if dataset has rows
if (nrow(PROMO_Data) > 0) {
  baseline_characteristics <- PROMO_Data %>%
    select(all_of(baseline_vars)) %>%
    tbl_summary(by = study_arm) %>%
    add_p() %>%
    modify_header(label = "**Variable**") %>%
    bold_labels()

  # Display only the formatted table
  baseline_characteristics
} else {
  cat("️ No data available for analysis.")
}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required libraries
library(dplyr)
library(gtsummary)

# Define malaria outcome variables
malaria_outcome_vars <- c(
  "total_malaria_episodes_during_pregnancy",
  "placental_malaria",
  "preterm_birth_ontoneo",
  "stillbirth",
  "infant_low_birth_weight_2500_g"
)

# Ensure that `study_arm` is in the dataset
if (!"study_arm" %in% colnames(PROMO_Data)) {
  stop("⚠️ Error: `study_arm` column not found in PROMO_Data. Please check column names.")
}

# Remove `stillbirth` and any other variable that has only one level
valid_outcomes <- malaria_outcome_vars[sapply(PROMO_Data[malaria_outcome_vars], function(x) length(unique(na.omit(x))) > 1)]

# Ensure `study_arm` is included in the selected variables
valid_outcomes <- c(valid_outcomes, "study_arm")

# Define label mappings dynamically based on valid variables
all_labels <- list(
  total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
  placental_malaria ~ "Placental Malaria (Histology)",
  preterm_birth_ontoneo ~ "Preterm Birth",
  stillbirth ~ "Stillbirth",
  infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
)

dynamic_labels <- all_labels[names(all_labels) %in% valid_outcomes]

# Define a custom function to format p-values
format_pvalue <- function(pvals) {
  sapply(pvals, function(x) {
    if (is.na(x)) {
      return("NA")  # Handle missing values properly
    } else if (is.numeric(x)) {
      if (x < 0.001) {
        return("< 0.001")  # Report very small values as "< 0.001"
      } else if (x < 0.01) {
        return(format(round(x, 3), nsmall = 3))  # Report between 0.001 and 0.01 to three decimal places
      } else {
        return(format(round(x, 2), nsmall = 2))  # Report >= 0.01 to two decimal places
      }
    } else {
      return("NA")  # Return "NA" for non-numeric values
    }
  })
}

# Generate table with formatted p-values
if (nrow(PROMO_Data) > 0 && length(valid_outcomes) > 1) {

  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(valid_outcomes)) %>%
    tbl_summary(
      by = "study_arm",
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",
        all_categorical() ~ "{n} ({p}%)"
      ),
      label = dynamic_labels
    ) %>%
    add_p(
      test = list(
        all_categorical() ~ "fisher.test",
        all_continuous() ~ "t.test"
      ),
      pvalue_fun = format_pvalue  # Apply the fixed p-value formatting function
    )

  # Print the updated table
  print(malaria_outcomes_table)

} else {
  message("⚠️ No data available for analysis or insufficient variability in variables.")
}

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}

# Load necessary packages
library(dplyr)
library(readr)
library(janitor)
library(tidyr)
library(here)
library(knitr)  # For table output

# ✅ Define the correct file path using `here` package
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"), show_col_types = FALSE)

# ✅ Clean Column Names
PROMO_Data <- PROMO_Data %>%
  clean_names()  # Standardizes column names (lowercase, removes spaces/special characters)

# ✅ Handle Missing Values
PROMO_Data <- PROMO_Data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%  # Replace NA in numeric columns with median
  mutate(across(where(is.character), ~ replace_na(., "Unknown")))  # Replace NA in categorical columns with "Unknown"

# ✅ Convert Data Types
PROMO_Data <- PROMO_Data %>%
  mutate(across(c(study_arm, delivery_mode, education_level), as.factor)) %>% # Convert categorical columns to factors
  mutate(across(contains("date"), as.Date, format="%Y-%m-%d"))  # Convert date columns

# ✅ Rename Key Variables for Clarity
PROMO_Data <- PROMO_Data %>%
  rename(
    `Study Arm` = study_arm,
    `Gestational Age at Enrollment (Weeks)` = gestational_age_at_enrollment_weeks,
    `Gestational Age at Delivery (Weeks)` = gestational_age_at_delivery,
    `Education Level` = education_level,
    `Detection of Malaria Parasites by Microscopy` = proportion_of_plasmodium_positive_blood_smears_during_pregnancy_by_microscopy,
    `Detection of Malaria Parasites by Microscopy or qPCR` = total_plasmodium_positive_routine_blood_smears_during_pregnancy
  )

# ✅ Remove Duplicate Rows
PROMO_Data_Cleaned <- PROMO_Data %>%
  distinct()  # Removes exact duplicate rows

# ✅ Standardize Categorical Variables with Proper Ordering
PROMO_Data_Cleaned <- PROMO_Data_Cleaned %>%
  mutate(
    # Age Categories
    Age_Category = factor(case_when(
      age_at_enrollment_years < 18 ~ "<18",
      age_at_enrollment_years >= 18 & age_at_enrollment_years <= 35 ~ "18–35",
      age_at_enrollment_years >= 36 & age_at_enrollment_years <= 55 ~ "36–55",
      age_at_enrollment_years > 55 ~ ">55",
      TRUE ~ "Unknown"
    ), levels = c("<18", "18–35", "36–55", ">55", "Unknown")),

    # Education Level Categories (With "Unknown" at the End)
    `Education Level` = factor(case_when(
      `Education Level` %in% c("A level", "O level", "None") ~ "Secondary/No Formal Education",
      `Education Level` == "Primary" ~ "Primary",
      `Education Level` %in% c("Tertiary", "University") ~ "Tertiary/Higher Education",
      is.na(`Education Level`) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("Primary", "Secondary/No Formal Education", "Tertiary/Higher Education", "Unknown")),

    # Gravidity (Number of Pregnancies) with Proper Ordering
    Gravidity = factor(case_when(
      gravidity == 1 ~ "1",
      gravidity == 2 ~ "2",
      gravidity >= 3 ~ "≥3",
      is.na(gravidity) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("1", "2", "≥3", "Unknown")),

    # Parity (Number of Births) with Proper Ordering
    Parity = factor(case_when(
      parity == 1 ~ "1",
      parity == 2 ~ "2",
      parity >= 3 ~ "≥3",
      is.na(parity) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("1", "2", "≥3", "Unknown"))
  )

# ✅ Create a Summary Table
summary_table <- PROMO_Data_Cleaned %>%
  summarise(
    `Total Participants` = n(),
    `Median Age` = median(age_at_enrollment_years, na.rm = TRUE),
    `Median Gestational Age at Enrollment` = median(`Gestational Age at Enrollment (Weeks)`, na.rm = TRUE),
    `Median Gestational Age at Delivery` = median(`Gestational Age at Delivery (Weeks)`, na.rm = TRUE)
  )

@@

Malaria is a life-threatening disease caused by Plasmodium parasites, which are transmitted to humans through the bite of an infected female Anopheles mosquito (Talapko et al., 2019). According to the World Health Organization, in 2022, approximately 249 million malaria cases were reported across 85 endemic countries—an increase of 5 million cases compared to 2021. Notably, Uganda alone contributed more than 597,000 cases during that period (WHO, 2023). Despite significant efforts in prevention and treatment, malaria remains a major global health challenge.

Vulnerable groups such as pregnant women, children, and immunocompromised individuals are at the highest risk, and Africa continues to bear the heaviest burden of this disease. Pregnant women with malaria tend to experience more severe clinical symptoms and poorer outcomes, including heightened risks of miscarriage, fetal loss, premature birth, low birth weight in newborns, and neonatal death (Caroline Lin Lin Chua et al., 2021). A multifaceted approach incorporating insecticide-treated bed nets (ITNs) and chemoprevention strategies has proven effective in mitigating the adverse health outcomes associated with malaria during pregnancy. Despite this progress, a 2017 report reveals that only 22% of pregnant women in Sub-Saharan Africa received an entire course of IPTp (Bauserman et al., 2019), underscoring a significant disparity in treatment access. Addressing this issue, the study seeks to: (a) examine the association between the frequency of malaria episodes during pregnancy and adverse birth outcomes, such as preterm birth and stillbirth; (b) assess whether infants born to mothers treated with DP are less likely to experience low birth weight compared to those treated with SP; and (c) explore the relationship between placental malaria and adverse birth outcomes, including preterm birth and low birth weight.

@@

Table 1: Baseline Characteristics and Pregnancy Outcomes of Pregnant Women
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}

# Load Required Packages
library(dplyr)
library(gtsummary)
library(janitor)
library(here)
library(knitr)

# ✅ Load Dataset (Ensure Correct File Path)
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"), show_col_types = FALSE)

# ✅ Clean Column Names
PROMO_Data <- PROMO_Data %>%
  clean_names()  # Standardizes names (lowercase, removes spaces/special characters)

# ✅ Handle Missing Values & Convert Data Types
PROMO_Data <- PROMO_Data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(where(is.character), ~ replace_na(., "Unknown"))) %>%
  mutate(across(c(study_arm, delivery_mode, education_level), as.factor)) %>%
  mutate(across(contains("date"), as.Date, format="%Y-%m-%d"))

# ✅ Rename Key Variables for Clarity
PROMO_Data <- PROMO_Data %>%
  rename(
    `Study Arm` = study_arm,
    `Gestational Age at Enrollment (Weeks)` = gestational_age_at_enrollment_weeks,
    `Gestational Age at Delivery (Weeks)` = gestational_age_at_delivery,
    `Education Level` = education_level,
    `Placental Malaria` = placental_malaria,
    `Total Malaria Episodes During Pregnancy` = total_malaria_episodes_during_pregnancy
  )

# ✅ Remove Duplicate Rows
PROMO_Data_Cleaned <- PROMO_Data %>% distinct()

# ✅ Standardize Categorical Variables (Ensuring "Unknown" Last)
PROMO_Data_Cleaned <- PROMO_Data_Cleaned %>%
  mutate(
    # ✅ Age Category (Ensure Correct Order)
    Age_Category = factor(case_when(
      age_at_enrollment_years < 18 ~ "<18",
      age_at_enrollment_years >= 18 & age_at_enrollment_years <= 35 ~ "18–35",
      age_at_enrollment_years >= 36 & age_at_enrollment_years <= 55 ~ "36–55",
      age_at_enrollment_years > 55 ~ ">55 (Older Adults)",
      TRUE ~ "Unknown"
    ), levels = c("<18", "18–35", "36–55", ">55 (Older Adults)", "Unknown")),

    # ✅ Education Level (Ensure "Unknown" is Last)
    `Education Level` = factor(case_when(
      `Education Level` %in% c("A level", "O level", "None") ~ "Secondary/No Formal Education",
      `Education Level` == "Primary" ~ "Primary",
      `Education Level` %in% c("Tertiary", "University") ~ "Tertiary/Higher Education",
      is.na(`Education Level`) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("Primary", "Secondary/No Formal Education", "Tertiary/Higher Education", "Unknown")),

    # ✅ Gravidity (Ensure "Unknown" is Last)
    Gravidity = factor(case_when(
      gravidity == 1 ~ "1",
      gravidity == 2 ~ "2",
      gravidity >= 3 ~ "≥3",
      is.na(gravidity) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("1", "2", "≥3", "Unknown")),

    # ✅ Parity (Ensure "Unknown" is Last)
    Parity = factor(case_when(
      parity == 1 ~ "1",
      parity == 2 ~ "2",
      parity >= 3 ~ "≥3",
      is.na(parity) ~ "Unknown",
      TRUE ~ "Unknown"
    ), levels = c("1", "2", "≥3", "Unknown"))
  )

# ✅ Define Baseline Characteristics Variables (Reorganized for Logical Flow)
baseline_vars <- c(
  # **Maternal Information First**
  "Age_Category", 
  "Education Level",
  "Gravidity", 
  "Parity", 
  "Placental Malaria", 
  "Total Malaria Episodes During Pregnancy", 
  "delivery_mode",

  # **Pregnancy Outcomes**
  "preterm_birth", 
  "small_for_gestational_age", 
  "spontaneous_abortion", 
  "stillbirth", 
  "neonatal_death",

  # **Infant Information**
  "infant_low_birth_weight_2500_g", 
  "infant_sex6"
)

# ✅ Ensure Selected Variables Exist in Dataset
baseline_vars <- baseline_vars[baseline_vars %in% colnames(PROMO_Data_Cleaned)]

# ✅ Generate Baseline Characteristics Table
baseline_table <- PROMO_Data_Cleaned %>%
  select(all_of(baseline_vars), `Study Arm`) %>%
  tbl_summary(
    by = `Study Arm`,
    statistic = all_categorical() ~ "{n} ({p}%)",
    missing = "no"
  ) %>%
  modify_header(label = "Characteristics") %>%
  bold_labels()

# ✅ Display the Table for Quarto Word Output
baseline_table

@@

Figure 1. Distribution of Adverse Pregnancy Outcomes by Study Arm
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required packages
library(ggplot2)
library(dplyr)
library(tidyr)

# Ensure correct column name for Study Arm
study_arm_col <- "Study Arm"

# Define pregnancy outcomes
pregnancy_outcomes <- c("preterm_birth", "stillbirth", "neonatal_death")

# ✅ Reshape data for plotting
pregnancy_data <- PROMO_Data_Cleaned %>%
  select(all_of(study_arm_col), all_of(pregnancy_outcomes)) %>%
  pivot_longer(cols = all_of(pregnancy_outcomes), names_to = "Outcome", values_to = "Present") %>%
  filter(!is.na(Present)) %>%  # Remove missing values
  group_by(`Study Arm`, Outcome, Present) %>%
  summarise(Total = n(), .groups = "drop")

# ✅ Convert outcome values to categorical (if numeric)
pregnancy_data$Present <- factor(pregnancy_data$Present, levels = c(0, 1), labels = c("Absent", "Present"))

# ✅ Plot Pregnancy Outcomes by Study Arm
ggplot(pregnancy_data, aes(x = Outcome, y = Total, fill = `Study Arm`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Pregnancy Outcomes",
       x = "Pregnancy Outcome",
       y = "Total Count",
       fill = "Study Arm") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "pink")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
